---
title: 'Homework 1: Moneyball'
output:
  html_notebook: default
  html_document: default
---

### Management summary

*Describe the goal of the project, summarize the main steps and findings*

### 1. DATA EXPLORATION 

The dataset contains 2276 observations of 17 variables.  
Each record (row) represents a professional baseball team from the years 1871 to 2006 inclusive. Each record has the performance of the team for the given year, with all of the statistics adjusted to match the performance of a 162 game season.

```{r message=FALSE}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(gridExtra)
library(MASS)
df_raw_train = read_csv("moneyball-training-data.csv",progress = F) 
```

Description of the variables:

![moneyball_variables.png]()

**Note**  
1) index variable will be discarded from the analysis  
2) BATTING_H variable appears to be a sum of single base hits and the counts of hits from the other variables listing the base hits of the types 2B, 3B, HR - thus a new variable BATTING_1B will be introduced to separate the effect of singles hits, and BATTING_H will be discarded


Data quality checks
- Share of missing values
- Proportion of unique observations

```{r}
# Preprocessing steps
# Discard INDEX, remove shorten variable names
df_train = df_raw_train %>% select(-INDEX)
names(df_train) = gsub("TEAM_", "", names(df_train))
names(df_train) = gsub("BATTING", "BAT", names(df_train))
names(df_train) = gsub("BASERUN", "BRU", names(df_train))
names(df_train) = gsub("FIELDING", "FLD", names(df_train))
names(df_train) = gsub("PITCHING", "PCH", names(df_train))
names(df_train) = gsub("TARGET", "TRG", names(df_train))

# Introduce new variable and discard BAT_H
df_train = df_train %>%
  mutate(BAT_1B = BAT_H - BAT_2B - BAT_3B - BAT_HR) %>% 
  select(-BAT_H)

sort(100*colMeans(is.na(df_train)),decreasing = T)
```

**Replacing NA variables**  
- BATTING_HBP -  batters hit by pitch - 91.6% of NAs - assume NA is equal to zero   (CONSIDER KICKING OUT COLUMN)
- BASERUN_CS - caught stealing - 33.9% of NAs - assume NA is equal to zero  - 
- FIELDING_DP - double plays - 12.6% of NAs - assume NA is equal to zero  
- BASERUN_SB -  stolen bases - 5.8% of NAs - assume NA is equal to zero  
- BATTING_SO - strikeouts by batters - 4.5% of NAs - assume NA is equal to zero  
- PITCHING_SO - strikeouts by pitchers - 4.5% of NAs - assume NA is equal to zero  
  
```{r}
# Discard BATTING_HBP
# Replace NA with mean value of non-empty records of a column

df_train = df_train %>% 
  select(-BAT_HBP) %>% 
  mutate_all(funs(ifelse(is.na(.),mean(.,na.rm = T),.)))
```


Variable summaries

```{r}
summary(df_train)
```

```{r}
sort(names(df_train))
```


Overview for each variable

```{r}

plot_overview = function(df, x) {
df = data.frame(df)
y = df[,x]

p1 = ggplot(df_train, aes(x= x, y = y)) + geom_boxplot() + xlab(x) + ylab("value")  
p2 = ggplot() + geom_density(aes(x=y),na.rm = T) + 
  geom_vline(aes(xintercept=mean(y,na.rm = T)), color = "red",linetype = 2) + xlab(x)
p3 = ggplot() + geom_point(aes(sample=y), stat = "qq", size = 0.1)
p4 = ggplot()  + geom_density(aes(x=log(y)),na.rm = T) + xlab(paste0("log(",x,")"))

grid.arrange(p1, p2, p3, p4, nrow = 2)
}

df = df_train
plot_overview(df,"TRG_WINS")

```

Target - almost normal

```{r}
plot_overview(df,"BAT_1B")
```
BAT_1B - left-skewed, a lot of outliers with high values, and two with very low

```{r}
plot_overview(df,"BAT_2B")
```
BAT_2B - nearly normal, with an each an extremely low and high outlier

```{r}
plot_overview(df,"BAT_3B")
```
BAT_3B - right-skewed, many extremely high values (try log transform - but has three zero values)

```{r}
plot_overview(df,"BAT_BB")
```
BAT_BB - has a mostly normal part and a weird very low tail (consider binning)

```{r}
plot_overview(df,"BAT_HBP")
```
BAT_HBP - nearly normal when exists, but most values are missing

```{r}


trg_var_name = "TRG_WINS"
x = "BAT_HBP"

plot_target_cor = function(df,x,trg_var_name){

df_cc = as.data.frame(df[complete.cases(df),]) 
trg_cor = cor.test(df_cc[,x],df_cc[,trg_var_name])
cor_out = paste("cor:",round(trg_cor$estimate,3),"p-val:" ,round(trg_cor$p.value,3))

x_pos = quantile(df_cc[,x],0.97)
y_pos = quantile(df_cc[,trg_var_name],0.001)
p = ggplot(data = df_cc, aes_string(x=x,y=trg_var_name)) +
  geom_point(size=1, alpha = 0.5) + geom_smooth(method = "lm") +
  annotate("text", x = x_pos, y = y_pos,label = cor_out) +
  ylab(trg_var_name)
p
}
```

```{r}
x = "BAT_BB"
trg_var_name = "TRG_WINS"


grid.arrange(plot_overview(df,x), plot_target_cor(df,x,trg_var_name),nrow = 2,heights=c(6,6))
```


IDEA:
1) Check outliers for validity
2) Discretize variables with a lot of outliers into quantiles (this way all outliers should fall into 1 bucket and lose leverage)

Pairwise plots to visualize distributions and inter-relationships

```{r}
# install.packages("GGally")
library("GGally")

ggpairs(select(df_train, matches("TRG_|BAT_")),
        title="Pairwise plots for the target and batting variables",
        upper = list(continuous = "cor"),
        lower= list(continuous = wrap("points", alpha = 0.3)))
```

```{r}
ggpairs(select(df_train, matches("TRG_|BRU_|FLD_")),
        title="Pairwise plots for the target and baserun/fielding variables",
        upper = list(continuous = "cor"),
        lower= list(continuous = wrap("points", alpha = 0.3)))
```

```{r}
ggpairs(select(df_train, matches("TRG_|PCH_")),
        title="Pairwise plots for the target and pitching variables",
        lower= list(continuous = wrap("points", alpha = 0.3)))
```

Correlation matrix

```{r}
ggcorr(df_train, nbreaks = 6,legend.size = 12,layout.exp = 2) + 
  ggtitle("Pairwise correlations")
```
Note that apart from BATTING_H and BATTING_HR there are no at least moderately strong positive or negative correlations with the target variable.

### 2. DATA PREPARATION 

fill from the above

### 3. BUILD MODELS 

```{r}
library(MASS)
fit <- lm(TRG_WINS~.,data=df_train)
step <- stepAIC(fit, direction="both")
step$anova # display results

```

```{r}
fit1 = lm(TRG_WINS ~ BAT_2B + BAT_3B + BAT_HR + BAT_BB + BAT_SO + BRU_SB + 
    PCH_H + PCH_SO + FLD_E + FLD_DP + BAT_1B, data = df_train)

summary(fit1)
plot(fit1)
fit1$model[2136,]
```


### 4. SELECT MODELS


Reference
LMR, MARR books
http://marcoghislanzoni.com/blog/2013/10/15/identifying-outliers-in-r-with-ggplot2/ 
